Accounting for uncertainty in simulating greenhouse gas emissions from dairy cattle in Kenya

### 2022-06-09 Manuscript_figures.Rmd
new rmd for the manuscript

---
title: "Accounting for uncertainty in simulation greenhouse gas emissions from dairy cattle in Kenya"
author: "Eike Luedeling, Cory Whitney, Todd Rosenstock"
bibliography:
 - references/packages.bib
 - references/papers.bib
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

@@

#install and load packages
# devtools::install_github("CWWhitney/uncertainty")
library(DiagrammeR)
library(decisionSupport)
library(ggplot2)
library(kableExtra)
library(magrittr)
library(plyr)
library(rmarkdown)
library(uncertainty)
library(chillR)
library(raster)
library(reshape2)

#Automatically write R package citation entries to a .bib file
knitr::write_bib(c(.packages(), 
                   'DiagrammeR',
                   'decisionSupport',
                   'ggplot2',
                   'kableExtra',
                   'magrittr',
                   'plyr',
                   'rmarkdown',
                   'uncertainty',
                   'chillR',
                   'raster',
                   'MASS'), 'references/packages.bib')
#resolves an issue with the scientific style of numbers,answer from @Paul Hiemstra: https://stackoverflow.com/a/25947542/4061993 
#from the documentation of ?options:
options(scipen=999)
```

```{r, warning=FALSE, eval=FALSE}
devtools::install_github("CWWhitney/uncertainty")
library(DiagrammeR)
library(decisionSupport)
library(ggplot2)
library(kableExtra)
library(magrittr)
library(plyr)
library(rmarkdown)
library(uncertainty)
library(chillR)
library(raster)
library(reshape2)
```

@@

```{r, warning=FALSE,eval=FALSE}
GHG_emissions<-mcSimulation(estimate=estimate_read_csv(input_table),
                            model_function=ghg_emissions,
                            numberOfModelRuns=1000,
                            functionSyntax="plainNames")
```

@@

@@

Here's the resulting file (only the first 3 farms are shown, and only the first 20 rows):

```{r}
read.table("data/farm_scenarios.csv", sep = ",", fileEncoding = "UTF-8-BOM",header=TRUE)[1:20,1:5]  %>%
    kableExtra::kbl(digits = 2)  %>%
    kableExtra::kable_classic_2(full_width = F)

@@

GHG_simulation_scenarios<-scenario_mc(base_estimate = estimate_read_csv("data/livestock_ghg_input_table.csv"),
                                 scenarios = read.csv("data/farm_scenarios.csv",
                                                      fileEncoding = "UTF-8-BOM"),
                                 model_function = ghg_emissions,
                                 numberOfModelRuns = 10,
                                 functionSyntax = "plainNames")

@@

```{r, eval=FALSE, include=FALSE}
dir.create("results")
save_temperature_scenarios(GHG_simulation_scenarios,"results","Baseline")
```

@@

Given these results we can now plot distributions of the various outputs from the model, using the `plot_distributions` function.

```{r plot_distributions, fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
# enteric_CH4=enteric_CH4,
enteric_CH4_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation_scenarios,
                   vars = "enteric_CH4",
                   method = "boxplot_density",
                   old_names = "enteric_CH4",
                   new_names = "CH[4]~Emissions~from~Enteric~Fermentation~(kg~CH[4])") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

enteric_CH4_dist_plot <- enteric_CH4_dist_plot + facet_wrap(. ~ name, labeller = label_parsed)

# milk_yield=sum(total_milk),
milk_yield_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation_scenarios,
                   vars = "milk_yield",
                   method = "boxplot_density",
                   old_names = "milk_yield",
                   new_names = "Total~Milk~Yield~(kg)") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

milk_yield_dist_plot <- milk_yield_dist_plot + facet_wrap(. ~ name, labeller = label_parsed)

# enteric_CH4_CO2eq=sum(enteric_CH4_CO2eq),
enteric_CH4_CO2eq_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation_scenarios,
                   vars = "enteric_CH4_CO2eq",
                   method = "boxplot_density",
                   old_names = "enteric_CH4_CO2eq",
                   new_names = "Enteric~CH[4]~Emissions~from~Livestock~(kg~CO[2]*'-eq')") + 
  theme(axis.title.x=element_blank())

enteric_CH4_CO2eq_dist_plot <- enteric_CH4_CO2eq_dist_plot + facet_wrap(. ~ name, labeller = label_parsed)

# mm_CH4_CO2eq=sum(mm_CH4_CO2eq),
mm_CH4_CO2eq_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation_scenarios,
                   vars = "mm_CH4_CO2eq",
                   method = "boxplot_density",
                   old_names = "mm_CH4_CO2eq",
                   new_names = "CH[4]~Emissions~from~Livestock~Manure~(kg~CO[2]*'-eq')") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

mm_CH4_CO2eq_dist_plot <- mm_CH4_CO2eq_dist_plot + facet_wrap(. ~ name, labeller = label_parsed)

# mm_N2O_CO2eq=sum(mm_N2O_CO2eq),
mm_N2O_CO2eq_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation_scenarios,
                   vars = "mm_N2O_CO2eq",
                   method = "boxplot_density",
                   old_names = "mm_N2O_CO2eq",
                   new_names = "N[2]*O~Emissions~from~Livestock~Manure~(kg~CO[2]*'-eq')") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())

mm_N2O_CO2eq_dist_plot <- mm_N2O_CO2eq_dist_plot + facet_wrap(. ~ name, labeller = label_parsed)

# feed_CO2=feed_CO2,
feed_CO2_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation_scenarios,
                   vars = "feed_CO2",
                   method = "boxplot_density",
                   old_names = "feed_CO2",
                   new_names = "CO[2]~Emissions~from~Livestock~Feed~Production~and~Transport~(kg)") + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())
feed_CO2_dist_plot <- feed_CO2_dist_plot + facet_wrap(. ~ name, labeller = label_parsed)

# on_farm=sum(on_farm)
on_farm_dist_plot <- plot_distributions(mcSimulation_object = GHG_simulation_scenarios,
                   vars = "on_farm",
                   method = "boxplot_density",
                   old_names = "on_farm",
                   new_names = "Total~Emissions~from~Livestock~(kg~CO[2]*'-eq')") + 
  theme(axis.title.y=element_blank())

on_farm_dist_plot <- on_farm_dist_plot + facet_wrap(. ~ name, labeller = label_parsed)

library(patchwork)
# remove the axis marks from the two lefthand-most plots and the x-axis title from the left and right plots

# plot with patchwork
layout <- "
           AAABBB
           CCCDDD
           EEEFFF
           GGGGGG
          "

enteric_CH4_dist_plot + milk_yield_dist_plot + 
  enteric_CH4_CO2eq_dist_plot + mm_CH4_CO2eq_dist_plot + 
  mm_N2O_CO2eq_dist_plot + feed_CO2_dist_plot + 
  on_farm_dist_plot + 
plot_layout(guides = "collect", design = layout)

ggsave("Fig_1_GHG_simulation_outputs.png",width=9,height=9)
```

# Possible errors resulting from input uncertainty

## Uncertainty at the farm scale

To illustrate the possible influence of input uncertainty on estimates of total farm emissions, we show the example of farm #5 from the household survey. This farm has 6 head of cattle, including 3 lactating cows. We simulate 1000 versions of this farm. We also do a precise calculation based on the exact numbers returned by the household survey and the default IPCC values.

### Total emissions of the farm

The following code and the resulting distribution estimate the total emissions of farm #5.

```{r, warning=FALSE}

@@

pd<-plot_distributions(mcSimulation_object = farm5_emissions,
                   vars = "on_farm",
                   method = "boxplot_density",
                   old_names = "on_farm",
                   new_names = "Total~Emissions~from~Livestock~(kg~CO[2]*'-eq')") + 
  theme(axis.title.y=element_blank())

pd + facet_wrap(. ~ name, labeller = label_parsed) + geom_vline(data=farm5_emissions_precise$y, aes(xintercept=on_farm),color="blue", lwd=3)

ggsave("Fig_2_uncertainty_illustration.png",width=5,height=3)

@@

The light blue distribution illustrates the estimated emissions from farm #5, considering uncertainty about all input variables. The blue line represents a classic best-bet estimate based on default values in the IPCC guidelines and some context-specific reasonable guesses.

### Emissions intensity

Naturally, the total emissions of the farm depend strongly on the size of the farm. A better indicator of the carbon footprint of milk production is the emissions intensity, i.e. the emissions per quantity of milk produced.

```{r, warning=FALSE, message=FALSE}
farm5_emissions$y[,"CO2eq_per_milk"]<-farm5_emissions$y$pc_on_farm/farm5_emissions$y$pc_milk_yield

histo<-ggplot(data=farm5_emissions$y, aes(CO2eq_per_milk)) +
geom_histogram(aes(y=..density..), colour="black", fill="white")+
 geom_density(alpha=.2, fill="#FF6666") +
  ylab("Relative probability") +
    xlab(expression(Emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^-1)))

histo + geom_vline(data=farm5_emissions_precise$y, aes(xintercept=CO2eq_per_milk),color="blue", lwd=3) + theme_bw()

ggsave("Fig_3_emissions_intensity_farm5.png",width=5,height=3)

@@

Also here, the blue line represents a classic best-bet estimate based on default values in the IPCC guidelines and some context-specific reasonable guesses.

## Population-wide emissions intensity of milk production

Since the emissions intensity of each household thus involves considerable uncertainty, such uncertainty should also be expected at the population level.

We can illustrate this uncertainty using a scatter plot. The following figure shows all combinations of total milk yield and total greenhouse gas emissions across all households. Total emissions are the sum of methane emissions from enteric fermentation and livestock manure, nitrous oxide emissions from manure and carbon dioxide emissions from livestock feed production and transport.

```{r, warning=FALSE, message=FALSE}
plotdata<-data.frame(in_var=GHG_simulation_scenarios$y$pc_milk_yield,
                     out_var=GHG_simulation_scenarios$y$pc_on_farm)

plot1 <- ggplot(plotdata, aes(x = in_var, y = out_var)) + 
  geom_point(shape = 20, color = "blue", size = 2) + 
  stat_smooth(method = "loessm", fullrange = TRUE) +
  geom_density_2d() +
  geom_rug(color="lightskyblue3") + 
  labs(y=expression(Total~emissions~(kg~CO[2]*'-eq'~head^{-1}~year^{-1})),
       x=expression(Total~milk~yield~(kg~head^{-1}~year^{-1}))) +
  theme_bw() +
  theme(legend.position = c(0.15, 0.9))

dens1 <- ggplot(plotdata, aes(x = in_var)) + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, fill = "white") +
  geom_density(fill="lightskyblue3") + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, alpha=0) + 
  theme_void() + 
  theme(legend.position = "none")

dens2 <- ggplot(plotdata, aes(x = out_var)) + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, fill = "white") +
  geom_density(fill="lightskyblue3") + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, alpha=0) + 
  theme_void() + 
  theme(legend.position = "none") + 
  coord_flip()

dens1 + plot_spacer() + plot1 + dens2 + 
  plot_layout(ncol = 2, nrow = 2, widths = c(4, 1), heights = c(1, 4))

ggsave("Fig_4_scatter_per_head_milk_emissions.png",width=5,height=3)

## Emissions intensity of milk production

As mentioned above, an indicator that is often of interest is the ratio between total emissions and the milk yield. We'll calculate this next. We then have to remove all households that don't produce any milk, because this ratio would be infinity for them. In fact, we'll remove all households that reported very low milk yields of <500 kg per year per capita, since these are unlikely to represent regularly operating dairy operations. For instance, they may be mostly focused on beef cattle production, or they may only have a small number of dairy cows that happened to be unproductive at the time of the survey.

```{r}
GHG_simulation_scenarios$y[,"CO2eq_per_milk"]<-GHG_simulation_scenarios$y$pc_on_farm/GHG_simulation_scenarios$y$pc_milk_yield

normal_milk_yield<-which(GHG_simulation_scenarios$y$pc_milk_yield>500)

Now we can plot the emissions intensity vs. the milk yield.

```{r, warning=FALSE, message=FALSE}

plotdata<-data.frame(in_var=dairy_households$y$pc_milk_yield,
                     out_var=dairy_households$y$CO2eq_per_milk,
                     Group="just one")

plot1 <- ggplot(plotdata, aes(x = in_var, y = out_var)) + 
  geom_point(shape = 20, color = "blue", size = 2) + 
  stat_smooth(method = "loess", fullrange = TRUE) +
  geom_density_2d() +
  geom_rug(color="lightskyblue3") + 
  labs(y=expression(Total~emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Total~milk~yield~(kg~head^{-1}~year^{-1}))) +
  theme_bw() +
  theme(legend.position = c(0.15, 0.9))

dens1 <- ggplot(plotdata, aes(x = in_var)) + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, fill = "white") +
  geom_density(fill="lightskyblue3") + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, alpha=0) + 
  theme_void() + 
  theme(legend.position = "none")

dens2 <- ggplot(plotdata, aes(x = out_var)) + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, fill = "white") +
  geom_density(fill="lightskyblue3") + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, alpha=0) + 
  theme_void() + 
  theme(legend.position = "none") + 
  coord_flip()

dens1 + plot_spacer() + plot1 + dens2 + 
  plot_layout(ncol = 2, nrow = 2, widths = c(4, 1), heights = c(1, 4))

ggsave("Fig_5_scatter_per_head_milk_emissions_intensity.png",width=7,height=7)

This can also be illustrated by a density surface.

```{r varkernel}

ggplot(plotdata, aes(x = in_var, y = out_var)) +
  stat_density_2d(
    geom = "raster",
    aes(fill = after_stat(density)),
    contour = FALSE
    ) +
  scale_fill_viridis_c() + 
  labs(y=expression(Total~emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Total~milk~yield~(kg~head^{-1}~year^{-1}))) +
  theme_bw()

ggsave("Fig_6_surface_per_head_milk_emissions_intensity.png",width=7,height=7)

The milk yield per capita is sometimes proposed as an indicator of greenhouse gas emissions. We can now examine the merits of this indicator. Essentially, its usefulness hinges on its relationship with greenhouse gas emissions intensity. We can investigate this by taking slices through the density surface.

# Milk yield as an indicator of GHG emissions intensity

## Probability of GHG emissions for a given milk yield

The following plot shows the emissions intensity that corresponds to a milk yield of 2000 kg per capita per year, based on the population of households covered in the survey.

```{r varkernelslicerange, message=FALSE, warning=FALSE}
slice<-uncertainty::varslice_resample(in_var = dairy_households$y$pc_milk_yield,
                                 out_var = dairy_households$y$CO2eq_per_milk, 
                                 expectedin_var=2000,
                           n = 1000,
                           n_samples = 1000,
                           out_var_sampling = 1000)

ggplot(data=data.frame(CO2eq_per_milk=slice$resampled), aes(CO2eq_per_milk)) +
geom_histogram(aes(y=..density..), colour="black", fill="white")+
 geom_density(alpha=.2, fill="#FF6666") +
  ylab("Relative probability") +
    xlab(expression(Emission~intensity~of~milk~production~at~2000~kg~milk~per~capita~(kg~CO[2]*'-eq'~kg^-1))) + theme_bw()

ggsave("Fig_7_emissions_intensity_all_farms.png")

We see considerable uncertainty about the emissions at this particular level of production, and the previous figure shows that emissions of all other levels are similarly variable. This raises considerable doubts about the usefulness of milk yield as an indicator of emissions intensity - at least when this indicator is taken at face value, without considering this uncertainty. Yet, some information is contained even in such an imperfect indicator, since, as also shown in the previous figure, there is clearly a correlation between milk yield and emissions intensity. The following proposes a mechanism for quantifying the level of certainty, with which a change in milk yield can reasonably be associated with a change in emissions intensity.

## Probability of an increase in milk yield being associated with an increase in GHG emission intensity

To derive the level of certainty, with which a change in milk yield is associated with an increase or decrease in emissions intensity, we have to investigate the probability distributions (as shown in the previous figure) of emissions intensity at two levels of milk production. This can be achieved by using the `varslice_resample` function for both production levels.

Let's look at this for production of 2000 and 3000 kg milk per capita per year. We'll take 1000 random draws of the emissions intensities that are associated with both levels and compare values among these samples.

```{r}
slice_2000<-uncertainty::varslice_resample(in_var = dairy_households$y$pc_milk_yield,
                                 out_var = dairy_households$y$CO2eq_per_milk, 
                                 expectedin_var=2000,
                           n = 1000,
                           n_samples = 1000,
                           out_var_sampling = 1000)
slice_3000<-uncertainty::varslice_resample(in_var = dairy_households$y$pc_milk_yield,
                                 out_var = dairy_households$y$CO2eq_per_milk, 
                                 expectedin_var=3000,
                           n = 1000,
                           n_samples = 1000,
                           out_var_sampling = 1000)

ggplot(data=slices, aes(Values, fill=Level)) +
 geom_density(alpha=.2) +
  ylab("Relative probability") +
    xlab(expression(Emissions~intensity~of~milk~production~at~specific~milk~yield~per~capita~(kg~CO[2]*'-eq'~kg^-1))) + labs(fill='Production level') + theme_bw()

ggsave("Fig_8_emissions_intensity_all_farms.png", width=7,height=5)

# sum(slice_3000$resampled < slice_2000$resampled) / length(slice_2000$resampled)

Now we want to determine the probability of emissions intensity at a production level of 3000 kg milk per capita and year being lower than emissions intensity at 2000 kg per capita and year. We can do this empirically by comparing the two sets of 1000 random values that are assembled in the probability distribution slices representing per capita milk production levels of 3000 and 2000 kg. We test for each index number (1-1000) if the respective element of the 3000-kg slice is smaller than the respective element of the 2000-kg slice. By adding up the results of all these comparisons (with TRUE coded as 1 and FALSE as 0) and dividing by 1000 (and optionally multiplying by 100% for a percentage), we obtain the probability that emissions intensity is really lower at the higher milk production level.

In this case, this comparison reveals that in `r sum(slice_3000$resampled < slice_2000$resampled) / length(slice_2000$resampled)*100` % of cases, emissions intensity at the higher milk production level is lower than at the lower level.

We can now generalize this result by doing similar calculations for multiple production levels and illustrating them for the whole population. We'll base this on slices representing production levels between 100 and 10,000 kg per capita and year, at 100-kg intervals.

```{r}

in_var = dairy_households$y$pc_milk_yield
out_var = dairy_households$y$CO2eq_per_milk

milk_yields_to_sample<-seq(100,10000,100)
milk_yields_to_sample<-milk_yields_to_sample[which(milk_yields_to_sample<max(in_var))]
milk_yields_to_sample<-milk_yields_to_sample[which(milk_yields_to_sample>min(in_var))]
slices<-list()

@@

for(i in 1:length(milk_yields_to_sample))
{slices[[i]]<-
  varslice_resample(in_var,
                      out_var,
                      expectedin_var = milk_yields_to_sample[i],
                      n_samples = 100000,
  out_var_sampling = 1000)$resample
 }

@@

#decreased_emissions_probability[,"yield_increase"]<-decreased_emissions_probability$new_yield-decreased_emissions_probability$Base_yield

@@

ggsave("Fig_9_probability_of_lower_emissions_intensity.png", width=7,height=5)

@@

The results indicate that whenever milk yields are higher on a particular farm than on another one, the probability that emissions intensity is lower is usually greater than 50% (slightly lower values are due to randomness in the resampling procedure), but that large differences are needed to be sure of this. The 90% confidence contour (shown in the plot as a black line) indicates the minimum level of milk yield that would be needed to state with 90% certainty that the emissions intensity is lower on the higher-productivity farm than on the respective lower-productivity one.

# Methodology for testing interventions

Farms may try to reduce their emissions intensity by making changes to their herd structure, feeding strategy or management practices. Programs aiming to effect emissions reductions in dairy production may want to reward such changes, and they may be looking to the milk production per capita as an indicator for the effectiveness of such measures. We now want to explore whether this is a reasonable strategy.

## Interventions

We will explore the impacts of three interventions that farmers may consider. We'll restrict this assessment to interventions we can easily implement in our modeling framework.

### Breed adjustment

We evaluate milk production for all the breeds found in the survey dataset. All cows of low-productivity breeds are then replaced by high-productivity cows.

### Herd structure intervention 1 - retiring unproductive bulls

All bulls that are beyond their productive life span are removed from the herd. We implement this by removing all animals of animal type 1, which denotes bulls >36 months of age.

### Herd structure intervention 2 - fewer unproductive replacement males

Many farms have more male calves and young bulls than are needed to meet the objectives of dairy production. Based on a real-life scenario, we target a reduction in this number by 23%. For this, we randomly select 23% of the farms and set the number of replacement males (animal type 3) to 0.

## Analysis strategy

In a model, we can simulate the implications of specific interventions by leaving all variables unchanged, except those that are modified by the intervention. To implement this using the decisionSupport package, we have to disable randomness for the variables we don't want to modify. For this, we can create a function (`run_model`) that simply applies our model to a set of values we specify. We can use this function to exactly reproduce the results of a model run. Note that this only works, because our model is perfectly deterministic. All variation in model outputs results from variation in inputs; the model itself doesn't include additional random processes.

We can now generate the outputs of a Monte Carlo model for our baseline situation (without applying interventions). The output object of the `mcSimulation` function doesn't only contain the output variables, however, but also a complete list of all random inputs for all model runs. As mentioned above, we could now apply the `run_model` function to these inputs (the `x` element of the `mcSimulation` result) to reproduce the outputs of the simulation. More importantly, we can now modify certain variables included in this `x` element to simulate interventions, and then run the model on these modified inputs to simulate the impacts of these interventions.

The following code contains the `run_model` function:

```{r}
run_model<-function(x_data,model)
  {
  varnames<-colnames(x_data)[which(!colnames(x_data)=="Scenario")]
  model_function_ <- function(x) {

    e <- as.list(sapply(X = varnames, FUN = function(i) as.numeric(unlist(x[i]))))
    eval(expr = body(model), envir = e)
  }
  y <- do.call(what = rbind, args = lapply(X = apply(X = x_data, 
                                                     MARGIN = 1, FUN = model_function_), FUN = unlist))
  returnObject <- list(y = data.frame(y), x = data.frame(x_data))
  returnObject$call <- match.call()
  class(returnObject) <- cbind("mcSimulation", class(returnObject))
  return(returnObject)
}

The following code can therefore precisely reproduce the results of the previous analysis:

```{r, eval=FALSE}
res<-run_model(GHG_simulation_scenarios$x,ghg_emissions)
```

To simulate interventions, we can now feed this function with modified x_data data.frames.

# Upgrading to higher-performance breeds

Replacing cows of low-productivity breeds with better-performing ones may reduce greenhouse gas emissions. We simulate an intervention that replaces animals of low-performance breeds with animals of high-performance breeds.

We first need to determine the performance and characteristics of each of the breeds that are present in the farm survey dataset (`Kenyaherds` dataset from above).

```{r, warning=FALSE}
cows<-Kenyaherds[which(Kenyaherds$animal_type %in% 4:6),]

>>>>

---
title: "Accounting for uncertainty in simulating greenhouse gas emissions from dairy cattle in Kenya"
author: "Eike Luedeling, Cory Whitney, Andreas Wilkes, Todd Rosenstock"
bibliography:
 - references/packages.bib
 - references/papers.bib
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

@@

#install and load packages
# devtools::install_github("CWWhitney/uncertainty")
library(DiagrammeR)
library(decisionSupport)
library(ggplot2)
library(kableExtra)
library(magrittr)
library(plyr)
library(rmarkdown)
library(uncertainty)
library(chillR)
library(raster)
library(reshape2)
library(patchwork)

knitr::write_bib(c(.packages(), 
                   'DiagrammeR',
                   'decisionSupport',
                   'ggplot2',
                   'kableExtra',
                   'magrittr',
                   'plyr',
                   'rmarkdown',
                   'uncertainty',
                   'chillR',
                   'raster',
                   'reshape2',
                   'patchwork'), 'references/packages.bib')
#resolves an issue with the scientific style of numbers,answer from @Paul Hiemstra: https://stackoverflow.com/a/25947542/4061993 
#from the documentation of ?options:
options(scipen=999)
```

```{r, warning=FALSE, eval=FALSE}
devtools::install_github("CWWhitney/uncertainty")
library(DiagrammeR)
library(decisionSupport)
library(ggplot2)
library(kableExtra)
library(magrittr)
library(plyr)
library(rmarkdown)
library(uncertainty)
library(chillR)
library(raster)
library(reshape2)
library(patchwork)
```

@@

```{r, warning=FALSE,eval=FALSE}
input_table <- "data/livestock_ghg_input_table.csv"
GHG_emissions<-mcSimulation(estimate=estimate_read_csv(input_table),
                            model_function=ghg_emissions,
                            numberOfModelRuns=1000,
                            functionSyntax="plainNames")
```

@@

read.table("data/farm_scenarios.csv", sep = ",", fileEncoding = "UTF-8-BOM",header=TRUE)[1:20,1:5]  %>%
    kableExtra::kbl(digits = 2)  %>%
    kableExtra::kable_classic_2(full_width = F)

@@

This is the resulting file (only the first 3 farms are shown, and only the first 20 rows):

@@

GHG_simulation_scenarios<-scenario_mc(base_estimate = estimate_read_csv("data/livestock_ghg_input_table.csv"),
                                 scenarios = read.csv("data/farm_scenarios.csv",
                                                      fileEncoding = "UTF-8-BOM"),
                                 model_function = ghg_emissions,
                                 numberOfModelRuns = 10,
                                 functionSyntax = "plainNames")
GHG_simulation_scenarios$y[,"CO2eq_per_milk"]<-GHG_simulation_scenarios$y$pc_on_farm/GHG_simulation_scenarios$y$pc_milk_yield
dir.create("results")
save_temperature_scenarios(GHG_simulation_scenarios,"results","Baseline")

@@

@@

Before we explore the results for the full population of farms, let's first look at one example. The following figure shows the greenhouse gas emissions, per farm and per head of cattle, for farm #5, which holds 12 head of cattle.

```{r, warning=FALSE, message=FALSE,fig.dim = c(8, 4)}

@@

farm5a<-ggplot(data=farm5_emissions$y, aes(on_farm)) +
geom_histogram(aes(y=..density..), colour="black", fill="white")+
 geom_density(alpha=.2, fill="#FF6666") +
  ylab("Relative probability") +
    xlab(expression(Total~emissions~from~livestock~(kg~CO[2]*'-eq'~year^-1))) + theme_bw() + ggtitle("a) Total emissions")

farm5_total <- farm5a + geom_vline(data=farm5_emissions_precise$y, aes(xintercept=on_farm),color="blue", lwd=3)

farm5_emissions$y[,"CO2eq_per_milk"]<-farm5_emissions$y$pc_on_farm/farm5_emissions$y$pc_milk_yield

farm5b<-ggplot(data=farm5_emissions$y, aes(CO2eq_per_milk)) +
geom_histogram(aes(y=..density..), colour="black", fill="white")+
 geom_density(alpha=.2, fill="#FF6666") +
  ylab("Relative probability") +
    xlab(expression(Emissions~intensity~(kg~CO[2]*'-eq'~kg^-1))) + ggtitle("b) Emissions intensity")

farm5_intensity<-farm5b + geom_vline(data=farm5_emissions_precise$y, aes(xintercept=CO2eq_per_milk),color="blue", lwd=3) + theme_bw()

farm5_total+farm5_intensity

ggsave("figures/Fig_1_emissions_from_farm_5.png",width=9,height=5)

@@

Now we look at the overall population.

```{r, warning=FALSE, message=FALSE, fig.dim = c(8, 4)}

all_farms_total<-ggplot(data=GHG_simulation_scenarios$y, aes(on_farm)) +
geom_histogram(aes(y=..density..), colour="black", fill="white")+
 geom_density(alpha=.2, fill="#FF6666") +
  ylab("Relative probability") +
    xlab(expression(Total~emissions~from~livestock~(kg~CO[2]*'-eq'~year^-1))) + theme_bw() + ggtitle("a) Total emissions")

all_farms_intensity<-ggplot(data=farm5_emissions$y, aes(CO2eq_per_milk)) +
geom_histogram(aes(y=..density..), colour="black", fill="white")+
 geom_density(alpha=.2, fill="#FF6666") +
  ylab("Relative probability") + theme_bw() +
    xlab(expression(Emissions~intensity~(kg~CO[2]*'-eq'~kg^-1))) + ggtitle("b) Emissions intensity")

all_farms_total+all_farms_intensity

ggsave("figures/Fig_2_emissions_from_all_farms.png",width=9,height=5)
```

We're interested in the suitability of total per capita milk production per year as an indicator of the greenhouse gas emissions intensity of milk production.

Here's the relationship between these two variables. We exclude farms with a per capita milk yield of <500 kg per year, since such farms can't be considered dairy farms and they would create outliers in many analysis.

```{r, warning=FALSE, message=FALSE, fig.dim = c(8, 6)}
normal_milk_yield<-which(GHG_simulation_scenarios$y$pc_milk_yield>500)

plot1 <- ggplot(dairy_households$y, aes(x = pc_milk_yield, y = CO2eq_per_milk)) + 
  geom_point(shape = 20, color = "blue", size = 2) + 
  stat_smooth(method = "loess", fullrange = TRUE) +
  geom_density_2d() +
  geom_rug(color="lightskyblue3") + 
  labs(y=expression(Emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Milk~yield~per~capita~(kg~head^{-1}~year^{-1})))+
  theme_bw() +
  theme(legend.position = c(0.15, 0.9))

dens1 <- ggplot(dairy_households$y, aes(x = pc_milk_yield)) + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, fill = "white") +
  geom_density(fill="lightskyblue3") + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, alpha=0) + 
  theme_void() + 
  theme(legend.position = "none")

dens2 <- ggplot(dairy_households$y, aes(x = CO2eq_per_milk)) + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, fill = "white") +
  geom_density(fill="lightskyblue3") + 
  geom_histogram(aes(y = ..density..),
                 colour = 1, alpha=0) + 
  theme_void() + 
  theme(legend.position = "none") + 
  coord_flip()

scatt<-dens1 + plot_spacer() + plot1 + dens2 + 
  plot_layout(ncol = 2, nrow = 2, widths = c(4, 1), heights = c(1, 4))

densi<-ggplot(dairy_households$y, aes(x = pc_milk_yield, y = CO2eq_per_milk)) +
  stat_density_2d(
    geom = "raster",
    aes(fill = after_stat(density)),
    contour = FALSE
    ) +
  scale_fill_viridis_c() + 
  labs(y=expression(Emissions~intensity~of~milk~production~(kg~CO[2]*'-eq'~kg^{-1})),
       x=expression(Milk~yield~per~capita~(kg~head^{-1}~year^{-1}))) +
  theme_bw()

dens1 + plot_spacer() + plot_spacer() + plot1 + dens2 + densi + 
  plot_layout(ncol = 3, nrow = 2, widths = c(4, 1), heights = c(1, 4))

ggsave("figures/Fig_3_scatter_and_surface_per_head_milk_emissions_intensity.png",width=10,height=6)
```

According to these distributions, the emissions intensity at a particular per capita milk yield can be described by a slice through the probability surface.

```{r varkernelslicerange, message=FALSE, warning=FALSE, fig.dim = c(12, 4)}
slice_2000<-uncertainty::varslice_resample(in_var = dairy_households$y$pc_milk_yield,
                                 out_var = dairy_households$y$CO2eq_per_milk, 
                                 expectedin_var=2000,
                           n = 1000,
                           n_samples = 1000,
                           out_var_sampling = 1000)
slice_3000<-uncertainty::varslice_resample(in_var = dairy_households$y$pc_milk_yield,
                                 out_var = dairy_households$y$CO2eq_per_milk, 
                                 expectedin_var=3000,
                           n = 1000,
                           n_samples = 1000,
                           out_var_sampling = 1000)

one_slice<-ggplot(data=data.frame(CO2eq_per_milk=slice_2000$resampled), aes(CO2eq_per_milk)) +
geom_histogram(aes(y=..density..), colour="black", fill="white")+
 geom_density(alpha=.2, fill="#FF6666") +
  ylab("Relative probability") +
    xlab(expression(Emissions~intensity~at~2000~kg~milk~per~capita~(kg~CO[2]*'-eq'~kg^-1))) + theme_bw() + ggtitle("a) Slice through the emissions intensity distribution")

two_slices<-ggplot(data=slices, aes(Values, fill=Level)) +
 geom_density(alpha=.2) +
  ylab("Relative probability") +
    xlab(expression(Emissions~intensity~at~specific~milk~yield~per~capita~(kg~CO[2]*'-eq'~kg^-1))) + labs(fill='Productivity level') + theme_bw() + ggtitle("b) Two slices at different productivity levels")

one_slice + two_slices

ggsave("figures/Fig_4_slices.png",width=13,height=6)

@@

By resampling from these distributions, we can determine the likelihood that a value of one of these distributions is greater than one of the other. We can systematically expand this to include more comparisons, in this case at intervals of 100 kg milk yield per capita.

```{r, fig.dim = c(8, 4)}

milk_yields_to_sample<-seq(100,10000,100)
milk_yields_to_sample<-milk_yields_to_sample[which(milk_yields_to_sample<max(dairy_households$y$pc_milk_yield))]
milk_yields_to_sample<-milk_yields_to_sample[which(milk_yields_to_sample>min(dairy_households$y$pc_milk_yield))]
slices<-list()

@@

for(i in 1:length(milk_yields_to_sample))
{slices[[i]]<-
  varslice_resample(dairy_households$y$pc_milk_yield,
                      dairy_households$y$CO2eq_per_milk,
                      expectedin_var = milk_yields_to_sample[i],
                      n_samples = 100000,
  out_var_sampling = 1000)$resample
 }

@@

@@

ggsave("figures/Fig_5_probability_of_lower_emissions_intensity.png", width=7,height=5)

@@

Now a description of the 5 interventions

Further figures to include:

- Fig. 6: Scatter plot that shows the impact of all interventions on per capita milk yield and emissions intensity
- Fig. 7: Scatter plot of all interventions against the backdrop of all samples drawn in second MC simulation (endline)
- Fig. 8: Intervention households shown on probability surface (Fig. 5)
- Fig. 9: Structural analysis - which household types have a chance of being detected
- Fig. 10: Errors (Type I and Type 2) - would be good to combine all interventions in one figure (might get crowded)
